---
title: "Risk, Odds, Ratios"
output: html_notebook
---


Setup
================================================================================

```{r}
#| label: setup
#| include: false

library(knitr)
library(tidyverse)
strokes <-
  read_csv("data/strokes.csv") |>
  filter(gender != "Other") |>
  mutate(stroke = as.logical(stroke))

```



Lab
================================================================================


Task 01
--------------------------------------------------------------------------------

- Question: The data includes some information about the employment of the
  patients. Which work type is the most prevalent in this data set? In other
  words, group by the work_type column and then count the number of each work
  type.
- Answer: `work_type`: Private

```{r}
strokes |> count(work_type) |> arrange(desc(n))
```


Task 02
--------------------------------------------------------------------------------

- Question: Use R to calculate the percent of patients who are children (based
  on work type/Task 01).
- Answer: 13.4

```{r}
strokes |> count(work_type) |> mutate(100*n/sum(n))
```


Task 03
--------------------------------------------------------------------------------

- Question: Which patients are more likely to have had a stroke, males or
  females?
- Answer: Male

```{r}

strokes |>
  group_by(gender) |>
  summarize(
    p = 100*mean(stroke)
  )

```


Task 04
--------------------------------------------------------------------------------

- Question: Do men with hypertension have a higher risk of stroke than women
  with hypertension?
- Answer: 

```{r}

strokes |>
  filter(hypertension == 1) |>
  group_by(gender) |>
  summarize(
    stroke = sum(stroke),
    n = n(),
    p = 100*stroke/n
  )

```


Task 05
--------------------------------------------------------------------------------

- Question: How many patients are over the age of 65?
- Answer:

```{r}
strokes |> filter(age > 65) |> count()
```


Task 06
--------------------------------------------------------------------------------

- Question: What is the risk ratio for stroke for men compared to women?
- Answer: 1.08

```{r}

risk <-
  strokes |>
  group_by(gender) |>
  summarize(
    stroke = sum(stroke),
    n = n(),
    risk = stroke/n
  )
risk

risk |>
  summarize(risk_ratio = risk[gender=="Male"]/risk[gender=="Female"])

```


Task 07
--------------------------------------------------------------------------------

- Question: What is the odds ratio for for stroke for men compared to women?
- Answer: 

- I'm not sure why, but I couldn't get the ! to help me calculate the
  denominator here.
- So I used n() - sum(stroke) to calculate the denominator.

```{r}

odds <-
  strokes |>
  group_by(gender) |>
  summarize(
    stroke = sum(stroke),
    not_stroke = n() - sum(stroke),
    odds = stroke/not_stroke
  )
odds

odds |>
  summarize(odds_ratio = odds[gender=="Male"]/odds[gender=="Female"])

```


Task 08
--------------------------------------------------------------------------------

Calculate the risk ratios for having hypertension and having heart disease.

- **Question:** Which chronic illness, hypertension or heart disease, increases
  the risk of having a stroke the most?
- **Answer:** 

```{r}



```


Task 09
--------------------------------------------------------------------------------

Thus far, we have the risk/odds of having a stroke for variables which have TWO
groups. Smoking status is different. Take a look at the following:

```{r}
strokes |> count(smoking_status)
```

This variable has four possibilities although Unknown isn't very
helpful. Calculate the odds ratio of smokers to those who have never
smoked. Then calculate the odds ratio of ex-smokers to those who have never
smoked.

- **Question:** Compared to those who never smoked, which group (smokers, former
  smokers) has the biggest risk of having a stroke according to this data?
- **Answer:**

```{r}



```
