---
title: "Lab"
output: html_notebook
---



Goals:

- Create a linear regression for total charges use `lm()`.
    - Evaluate how residuals differ when we perform the regression between total charges as compared to log of total charges.
- Create a linear regression for total charges using `rpart()` (a decision tree).
- Create a decision tree which can predict if a 

Things to consider:

- This is money related lab. Therefore, there are dollar signs in the write-up. Because I frequently have more than one in a single paragraph, I have to "escape" the dollar signs so markdown understands I want an ACTUAL dollar sign and that I'm not using it as a special command. This is why you will see things like, \$1,000,000 in the paragraphs below. If I don't do that, rendering to HTML would break in spectacular and hysterical ways.
- I recommend you complete the lab in RStudio before opening Canvas. Yes, this will involve some scrolling around in RStudio, but I think you will be less frustrated if you complete the tasks in RStudio before moving to Canvas.



# Setup

Remember that each R Session is run separately. When you leave RStudio, that R Session is shut down and goes away. The only thing which continues to exist is any on-disk data (or other side-effects) created during the session and the R Script (R Markdown here) written. This R code should enable you to reproduce the important/valuable parts of your previous R Session, but you must first run the script. If all you want is to re-run the `setup` chunk, you can easily do that by accessing the drop-down next to the `Run` button above and selecting `Run Setup Chunk`.

```{r setup, message=FALSE, warning=FALSE}
rm(list = ls())
library(knitr)
library(modelr)
library(rpart)
library(rpart.plot)
library(tidyverse)

## This prevents R from using scientific notation.
options(scipen=999)

## This imports the data from the CSV file in the data folder, and then adds
## the charge_group column on the end.
births <- 
  read_csv("data/births.csv") %>% 
  mutate(
    ## I use line breakes to make the code easier to read.
    charge_group = case_when(total_charges > 10000~"High Cost", TRUE~"Low Cost"),
    ## Decision trees tend to see numbers as continuous variables so we will convert
    ## gender into a string so rpart see it as a factor and not a continuous variable.
    gender = case_when(gender == 0~"Female", gender==1~"Male", TRUE~"Unk"),
  ) %>% 
  filter(!is.na(total_charges))

births
```

## About 

This data set was obtained from the Book, "Biostatistics for the Biological and Health Sciences" by Triola, Triola, and Roy. This is the "Births" data set which is comprised of 400 births from upstate New York, including Albany.

Columns:

- ID: Unique birth identifier.
- Facility: Name of the hospital where the birth occurred.
- Insurance: Name of the mother's insurance, which paid for the birth.
- Gender: Gender of the newborn.
    - 0 = Female
    - 1 = Male
- Length Of Stay: Duration of the hospital stay for the newborn in days.
- Admitted: Day of the week the mother was admitted to hospital.
- Discharged: Day of the week the infant was discharged from hospital.
- Birth Weight: Birth weight of the newborn in grams.
- Total Charges: Total cost of the birth in US dollars.



# EDA

Although the average birth costs over \$15,000, we can prove that the vast majority of births cost \$10,000 or less.

## Task 01

What is the average cost for a birth from this data set?

- Recall that `summarize()` will reduce a table from many rows to one.

```{r}
## YOUR CODE HERE!!!

```

## Task 02

As you can imagine, the cost of a birth varies, a lot. Draw a density plot of the `total_charges` column.

```{r}
## YOUR CODE HERE!!!

```

You should see a peak on the far left of the plot and a loooooong tail off to the right. The highest cost in this data set is over \$400,000! For this lab, we will define any cost over \$10,000 as a "High Cost" birth. Any birth that costs \$10,000 or less, we will call a "Low Cost" birth. The code below should look familiar to you. It uses our dplyr verbs to transform our births data (400 rows) into two rows using `group_by()` and `summarize()`.

```{r}
births %>% 
  ## Note - I am not using filter here!
  group_by(charge_group) %>% 
  summarize(
    n_births = n(),
    charges_births = sum(total_charges, na.rm = TRUE),
    charges_missing = sum(is.na(total_charges)),
    min_charges = min(total_charges, na.rm = TRUE),
    avg_charges = mean(total_charges, na.rm = TRUE),
    max_charges = max(total_charges, na.rm = TRUE)
  ) %>%
  mutate(
    p_births = round(n_births/sum(n_births)*100,2),
    p_charges_births = round(charges_births/sum(charges_births)*100,2)
  ) %>% 
  ## summarize/mutate blindly add the new columns to the end of the result set.
  ## Here, I wanted my returns in a specific order, so I am using select to 
  ## place columns next to one another in an order which is different than how
  ## the code runs internally. Basically, this just makes it look nicer.
  select(
    charge_group,
    n_births,
    p_births,
    min_charges,
    avg_charges,
    max_charges,
    charges_births,
    p_charges_births,
    charges_missing
  )
```

- Roughly 85% of our births account for merely 15% of our costs. While 15% of our births account for 85% of our costs.
    - This is ALMOST an 80/20 split. It is a power-law distribution.
    - [Wikipedia: Pareto Principle](https://en.wikipedia.org/wiki/Pareto_principle)
    - This may seem extreme (it is). This is also incredibly normal in real world data.
- Congratulations, you work at CDPHP or MVP or ACPHS Metro Insurance and you are tasked with reducing how much the plan pays for delivering babies.
    - Where are you going to focus your efforts?
- Are you going to focus your energy on the 85% of births or 15%?
    - You _could_ try to renegotiate the cost of this or that with Albany Med/St. Peters.
    - Routine deliveries have routine costs.
    - **As a payer, your annual expenditures are driven by the exceptional cases.**

The plot before, which is very similar to what you were supposed to write above, shows the distribution of charges across our two stratifications.

```{r}
births %>%
  ## We cannot plot the charges of the 25 low-cost births here, so I filtered
  ## them out to avoid a warning from ggplot.
  filter(!is.na(total_charges)) %>% 
  ggplot(aes(total_charges)) +
  geom_density() +
  facet_wrap(~charge_group, scales = "free")
```

As a payer you might try to identify pregnancies that are at high risk of being in this high-cost group. But, what are the risk factors for having an expensive baby?



# Linear Regression

A couple of weeks ago, we learned how to develop a linear regression using the `lm()` function. The following tasks should help you develop a reasonable linear model for our birth costs. Remember, a linear regression equation looks like:

$Y = a + bX$

(↑ This will look nicer if you hit the Preview button above this notebook.)

Above, we will treat Y as `total_charges`, a will be our intercept and bX will be the slope applied to some function of X. Also, recall that we can judge the accuracy of this model simply by measuring the distance between our predicted values and our actual values and calling these the residuals. We want to minimize our residuals and keep them normally distributed.

## Task 03

Draw a scatter plot of `total_charges` as a function of `length_of_stay`, stratified by `charge_group`.

In other words:
- Put `length_of_stay` on the X axis and `total_charges` on the y axis.
- Use `facet_wrap()` to draw one plot for High Cost births and one plot for Low Cost births.
- I recommend using the parameter `scale = "free"` in `facet_wrap()` in order to be able to better read/interpret the plots.
    - If you don't do this, your Low Cost plot will be basically unreadable.

```{r}
##YOUR CODE HERE!!!

```

In Canvas, I will ask you to upload the plot.

## Task 04

Draw another scatter plot, this time removing the stratification (`facet_wrap()`). Instead, I want you to color the scatter plot by `charge_group`. This should result in single plot, where there are points in two colors, one for Low Cost Births and another color for High Cost births.

In other words, just draw a scatter plot of total charges as a function of length of stay, but using color to highlight the differences between our two charge groups.

Please add the linear regression to this plot by using `geom_smooth(method = lm)`

```{r}
## YOUR CODE HERE!!!

```

In Canvas, I will not ask for this plot. Instead, I want to know if you believe the x-intercept, which is the value of y where x = 0, is greater than, less than, or equal to zero.

## Task 05

Write a linear model, called `lin_mod_1` using the `lm()` function we used a couple of weeks ago, to create a simple linear model of `total_charges` as a function of `length_of_stay`.

```{r}
## Uncomment the lines below and complete the code by replacing the ?.
## lin_mod_1 <- lm(total_charges~?, data = births)
## summary(lin_mod_1)
```

In Canvas, I will ask you to report two things from this model:

- What is the Adjusted R-Squared of this model? 
- Is the p-value less than .05 for the over-all model?

You should see that the Estimate for:

- Intercept: -5201.5
- length_of_stay: 4618.1

PLEASE look at the summary output until you locate these two bits of information. The intercept is telling us that when days = 0, the cost of a birth costs less than 0, but that is fine, because there is no such thing as an in-hospital delivery which takes zero days. The 4,618.1 value is telling us that, on average, for every additional day spent in the hospital, total charges go up by $4,618.10 dollars. Thus, on average, the cost of a delivery where the baby stays in the hospital for two days is:

$Y = -5,201.5 + 4618.1*2 $

And this is nothing more than the details of: 

$Y = a + b*X $

filled in. Again, if you want to see the LaTeX rendered, hit preview.

Now, run the code below and look at the results. Please see that there are two outputs. You have a plot, but there is also some text to look at.

```{r}
## Remember, we can assess our residuals!
lin_mod_1_results <- augment(lin_mod_1)
lin_mod_1_results
ggplot(data = lin_mod_1_results, aes(.resid)) + geom_density()
```

If you did everything right:
- Your average residual should be ALMOST ZERO.
- The standard deviation of your residuals should be (roughly) normally distributed, but there are going to be some extreme values.

Using this, we can calculate our residual standard error (which should have been 24310 above in `summary(lin_mod_1)`):

- Try to understand the formular here: https://www.statology.org/residual-standard-error-r/
- And then look at the code below.
- The summary function told us 24310 thanks to a little rounding.
- Reducing our residual standard error is similar to reducing the standard deviation of our residuals. The residual_standard_error does a better job of penalizing us for making the model more complex AND was easier to calculate in some ways before the invention of the computer.

```{r}
## Yes, summary does a little rounding.
lin_mod_1_results %>%
  summarize(
    sum_of_squares = sum(.resid^2),
    degrees_of_freedom = n() - 2,
    residual_standard_error = sqrt(sum_of_squares/degrees_of_freedom)
  )
```


## Task 06

Look at the other columns in `births`. In modeling, we call these columns "features". Chosen well, adding features to a model can improve it. I want you to look at the features you have available to you in `births` and choose one or two to add to your model. Call this new model `lin_mod_2`.

```{r}
## Complete the code below by removing the comment and replacing the ?.
## Warning, there may be another bug for you to have to fix in my code.
## lin_mod_2 <- lm(total_charges~length_of_stay+?, data = births)
## summary(lin_mode_2)
```

In Canvas, I am going to ask you to tell me the formula you chose and how you went about choosing the feature(s) you added to the model. Justify these additions by telling me how the affected the Adjusted r-squared of the model and the residuals of the model.


## Task 07

In class we have used CART models for classification. But they can also be used for regression analyses. Follow along with this code. And please take the time to read it and try to understand what it is doing and then answer the question below.

```{r}
lin_cart_mod_1 <- 
  rpart(total_charges~length_of_stay, data = births)
summary(lin_cart_mod_1)
rpart.plot(lin_cart_mod_1)
```

You will notice that this is a quite simple model, but it is a model.

- And we can use this model to make predictions.
- And we can assess the quality of those predictions.

```{r}
## Here, we are adding our predictions back into our original data set.
predicted <- add_predictions(births, lin_cart_mod_1)

## And here we are calculating the residuals.
## For a regular linear model, R does this for us. Here, we have to do it.
## Meh.
predicted <- 
  predicted %>% 
  mutate(resid = pred-total_charges)

predicted %>% 
  summarize(
    sum_of_squares = sum(resid^2),
    degrees_of_freedom = n() - 2,
    residual_standard_error = sqrt(sum_of_squares/degrees_of_freedom)
  )

ggplot(predicted, aes(resid)) + geom_density()
```

Using criteria such as residual standard error, which model performed better? 

- `lin_mod_1`
- `lin_cart_mod_1`

Don't confuse `lin_mod_1` which I wrote for you with `line_mod_2` which you wrote.


## Task 08

Can you build a BETTER CART model which outperforms our model above? Call this model `lin_cart_mod_2`. Is `charge_group` ever useful? In Canvas, tell me what your model is and why you chose it. Be sure to use some of the results I demoed for you above.

```{r}
## YOUR CODE HERE!!!

```
