---
title: "Ratios"
format: html
---



This lab is a little different from any previous lab. In this lab you will
calculate risk, odds, and risk/odds ratios. And you will do so using a data set
related to stroke.

Congrats. You are doing public health.

I also want you to use this lab to refresh your dplyr skills. I know it has been
a while since we really flexed our dplyr muscles.


Setup
================================================================================

```{r}
#| label: setup

library(knitr)
library(rio)
library(tidyverse)
strokes <-
  read_csv("data/strokes.csv") |>
  filter(gender != "Other")

```



Data - Strokes
================================================================================

Source: [Kaggle Stroke Prediction
Dataset](https://www.kaggle.com/fedesoriano/stroke-prediction-dataset)

> According to the World Health Organization (WHO) stroke is the 2nd leading
> cause of death globally, responsible for approximately 11% of total
> deaths. This dataset is used to predict whether a patient is likely to get
> stroke based on the input parameters like gender, age, various diseases, and
> smoking status. Each row in the data provides relavant information about the
> patient.

```{r}
strokes
```

1. id: unique patient identifier
2. gender: "Male", "Female" or "Other"
3. age: age of the patient
4. hypertension: 0 if the patient doesn't have hypertension, 1 if the patient
   has hypertension
5. heart_disease: 0 if the patient doesn't have any heart diseases, 1 if the
   patient has a heart disease
6. ever_married: "No" or "Yes"
7. work_type: "children", "Govt_jov", "Never_worked", "Private" or
   "Self-employed"
8. residence_type: "Rural" or "Urban"
9. avg_glucose_level: average glucose level in blood
10. bmi: body mass index
11. smoking_status: "formerly smoked", "never smoked", "smokes" or "Unknown"*
12. stroke: 1 if the patient had a stroke or 0 if not *Note: "Unknown" in
smoking_status means that the information is unavailable for this patient

There aren't any oddly named columns in this data.



Lab
================================================================================


Task 01
--------------------------------------------------------------------------------

- **Question:** The data includes some information about the employment of the
  patients. Which work type is the most prevalent in this data set? In other
  words, group by the work_type column and then count the number of each work
  type.
- **Answer:** 
- Helpful Hints:
    - To prevent confusion, your options below are identical to what you should
      have found in your data. In Canvas complete this task with one of the
      following:
          1. children
          2. Govt_job
          3. Never_worked
          4. Private
          5. Self-employed
      - The entries in the `work_type` column are formatted oddly, but that's
        OK.There's always _something_ funny about a data set.

```{r}

## YOUR CODE HERE!!!

```


Task 02
--------------------------------------------------------------------------------

- **Question:** Use R to calculate the percent of patients who are children
  (based on work type/Task 01).
- **Answer:** 
- Helpful Hints:
    - In other words, divide the work_type count by the total.
    - And yes, you CAN do this without R, but the point is to do this using R.
    - And the easiest way is to calculate the percent for all of the work types.

```{r}

## YOUR CODE HERE!!!

```


Task 03
--------------------------------------------------------------------------------

- **Question:** Which patients are more likely to have had a stroke, males or
  females?
- **Answer:**
- Helpful Hints:
    - We want to calculate the percentage of patients with a stroke, stratified
      by gender.

```{r}

## YOUR CODE HERE!

```


Task 04
--------------------------------------------------------------------------------

- **Question:** Do men with hypertension have a higher risk of stroke than women
  with hypertension?
- **Answer:** 
- Helpful Hints:
    - This is ALMOST the same question.
    - All we need to do is to filter down to those with hypertension.
    - Filter should come before group by.

```{r}

## YOUR CODE HERE!!!

```


Task 05
--------------------------------------------------------------------------------

- **Question:** How many patients are over the age of 65?
- **Answer:**

```{r}

## YOUR CODE HERE!

```


Task 06
--------------------------------------------------------------------------------

- **Question:** What is the risk ratio for stroke for men compared to women?
- **Answer:** 

Helpful Hint:

- It may help to calculate this all as two steps.
- First calculate the risk of stroke for men and women.
    - Save this in a table called risk.
- Then calculate the risk ratio.
- And remember, if you can't figure out the fancy code for the risk ratio but
  you can figure out the gender-specific risk, you can just use R like a
  calculator and calculate the risk ratio manually.

```{r}

## YOUR CODE HERE!!!

```


Task 07
--------------------------------------------------------------------------------

- **Question:** What is the odds ratio for for stroke for men compared to women?
- **Answer:** 

```{r}

## YOUR CODE HERE!

```

- PLEASE PLEASE PLEASE observe the odds nearly equals the risk and the odds
  ratio and risk ratio are nearly identical.
- This is because strokes are rare compared to drowning on the Titanic.
- That said, this data does seem to have an unusually high stroke rate.


Task 08
--------------------------------------------------------------------------------

Calculate the risk ratios for having hypertension and having heart disease.

- **Question:** Which chronic illness, hypertension or heart disease, increases
  the risk of having a stroke the most?
- **Answer:** 

```{r}

## YOUR CODE HERE!

```


Task 09
--------------------------------------------------------------------------------

Thus far, we have the risk/odds of having a stroke for variables which have TWO
groups. Smoking status is different. Take a look at the following:

```{r}
strokes |> count(smoking_status)
```

This variable has four possibilities although Unknown isn't very
helpful. Calculate the odds ratio of smokers to those who have never
smoked. Then calculate the odds ratio of ex-smokers to those who have never
smoked.

- **Question:** Compared to those who never smoked, which group (smokers, former
  smokers) has the biggest risk of having a stroke according to this data?
- **Answer:**

```{r}

## YOUR CODE HERE!

```
