---
title: "Risk, Odds, Ratios"
format: html
---



Goals
================================================================================

- Some funky base-R skillz
- Introduce you to my dear friends:
    - Risk
    - Odds
    - Risk Ratio
    - Odds Ratio

I know this seems like a lot of new ideas, but I'm really only introducing a two
new ideas:

1. Odds
2. Calculating a ratio

In the coming weeks, we will connect these ideas to logistic regression, but
_that_ is for another day.



Setup
================================================================================

This week's data consists of a data set we have seen before.

```{r}
#| label: setup
#| include: false

library(knitr)
library(janitor)
library(patchwork)
library(tidymodels)
library(tidyverse)

appointments <- read_csv("data/Appointments.csv") |> clean_names()

```



Data
================================================================================

- Do you remember the appointments data set from the midterm?
    - I am sympathetic if you chose to block it out.
- For the midterm, I altered a few spelling irregularities for you.
    - Today we will see it in its full glory.


```{r}
appointments |> head()
```

Data Dictionary
--------------------------------------------------------------------------------

  - patient_id: Unique identifier for that patient.
  - appointment_id: Unique identifier for each appointment
  - gender: Male (M) or Female (F).
  - appointment_date: The Date (YYYY-MM-DD) of the appointment
  - appointment_day: An abbreviation of the name of the day of the
    appointment (Mon, Tue, etc.)
  - appointment_month: The abbreviation of the name of the month of the
    appointment.
  - age: Patient age, on the day of the appointment, in years.
  - neighbourhood: Where the appointment takes place.
    - Note the unusual spelling of this column name.
  - scholarship: TRUE or FALSE
      - For more detail:
        <https://en.wikipedia.org/wiki/Bolsa_Fam%C3%ADlia>
  - hipertension: Does the patient have hypertension? TRUE or FALSE
      - Note the unusual spelling of this column name.
  - diabetes: Does the patient have diabetes? TRUE or FALSE
  - alcoholism: Does the patient have diabetes? TRUE or FALSE
  - handcap: Does the patient have a handicap? TRUE or FALSE
      - Note the unusual spelling of this column name.
      - This is not my typo.
  - received_sms: Did the patient receive a SMS? TRUE or FALSE
  - no_show: Did the patient no-show? TRUE or FALSE



Funky Base-R Skillz
================================================================================

- Every dataframe is composed of rows and columns.
- Every element in a column must be the same type.
- In R, a column can also be called a vector.
- A vector is just a collection of information with the same data type.

```{r}

foo <- c(1, 2, 3, 4)
bar <- c(TRUE, FALSE, TRUE, FALSE)
baz <- c("a", "B", "c", "D")
foo
bar
baz

```

- I have created THREE vectors, foo, bar, and baz.
- Each is epic in its own way.
- And we can operate on these.


```{r}
sum(foo)
```


```{r}
sum(bar)
```


```{r}
length(baz)
```

We can perform filtering operations on vectors using SQUARE BRACKET NOTATION.

```{r}
bar[foo < 3]
```

```{r}
baz[foo <= 3]
```

```{r}
foo[bar]
```


```{r}
baz[bar]
```

```{r}
baz[!bar]
```

- We filter our vectors by performing a boolean test.
- I know this seems random, but it will be important later.



Percentage, Risk, Odds
================================================================================

Because `no-show` can be treated as a 1/0 column, we can represent the 
probability of a no-show appointment as:

 $P(no\_show) = \frac{\sum_{n=1}^{n} no\_show}{n}$

The probability of a no-show case is the count of no-show cases divided by the
total number of cases.

Percent No Show
--------------------------------------------------------------------------------

- **Question:** What percentage of appointments no-show?
- **Answer:** 20.2%
- Helpful Hints:
    - I hope this feels like a review.

```{r}

# Three paths, one destination:

# One
prc_1 <-
  appointments |>
  summarize(
    den = n(),
    num = sum(no_show),
    p = round(100*num/den,1)
  )

# Two
prc_2 <-
  appointments |>
  summarize(
    den = n(),
    num = sum(no_show)
  ) |>
  mutate(
    p = round(100*num/den,1)
  )

# Three
prc_3 <-
  appointments |>
  summarize(
    den = n(),
    num = sum(no_show),
    p = round(100*mean(no_show),1)
  )


wrap_table(prc_1) / wrap_table(prc_2) / wrap_table(prc_3) 
```

Calculating a rate requires a numerator and a denominator. To calculate the
percentage of appointments which no-show can take three steps:

1. Denominator (den): Total number of appointments.
2. Numerator (num): Number of no-shows appointments.
3. Percentage (p): 100 * (Numerator/Denominator)

All three of these steps can take place in a single summarize command or you can
calculate the percentage in a separate mutate function as shown above.

If you use group_by, it can force you to use a mutate.


Risk No Show
--------------------------------------------------------------------------------

Now calculate the over-all RISK of a no-show appointment.

```{r}

# You can pick any way you want to calculate late this.
# As long as your numbers match my numbers.

rsk_1 <-
  appointments |>
  summarize(
    den = n(),
    num = sum(no_show),
    risk = round(100*num/den,1)
  )
rsk_1

```

- Risk is an important measure of the likelihood of a particular outcome you   
  are concerned with.
    - The only difference between a percentage calculation and a risk 
      calculation is the label.
    - Here, I am expressing risk as a percentage.
    - You can also express it as a simple rate. (Not multiplied by 100)
    - Or you can express risk as a ratio of some other base such as 1,000.
- More formally: The ratio of the number of events that produce the outcome of
  interest to the total number of events.
    - You don't have to multiply by 100. We often do, but it isn't required.
    - But you do want to be clear if you have or not.
    - In other words: .202 risk is the same as 20.2% risk. Just make sure your
      audience knows if you are using the risk rate or the percent risk.

Congratulations - YOU ALREADY KNEW HOW TO DO THAT!!!


Odds No Show
--------------------------------------------------------------------------------

- The odds of a no-show are a related concept.
- Odds provide another measure of the likelihood of a particular outcome.
- The ratio of the number of events that produce the outcome of interest to the
  number that do not.
- This ratio for odds is always higher than the risk.


```{r}
# These two methods are functionally equivalent.


odd_1 <-
  appointments |>
  summarize(
    den = n() - sum(no_show),
    num = sum(no_show),
    odds = num/den
  )

# OR

odd_2 <-
  appointments |>
  summarize(
    den = sum(!no_show),
    num = sum(no_show),
    odds = num/den
  )

wrap_table(odd_1) / wrap_table(odd_2)

```

- Never multiply odds by 100.
    - BECAUSE IT IS NOT A PERCENTAGE.
    - **I can't stress this enough.**
- Odds are often written as a fraction or ratio.
  - In this example, the odds are _approximately_ 1 to 4.
  - Read as: For every one no-show, you would expect about four arrived
    appointments.
  - For rare events, the odds and the risk are very similar (converge).
      - See below for a demo of how the risk roughly equals odds for adequately
        rare events.
  - When expressed as a decimal, the odds tend to "over-dramatize" likelihood of
    an event because we more naturally think in terms of risk than odds.

Comparing Hypothetical Odds to Risk:

| No Show | Arrived | Appointments |  Risk |  Odds |
|--------:|--------:|-------------:|------:|------:|
|      20 |      80 |          100 | .2000 | .2500 |
|      20 |     800 |          820 | .0240 | .0250 |
|      20 |    8000 |         8020 | .0020 | .0025 | 

- The above table contains the risk ratio, not the percent.
- This makes it easier to compare to the odds.
- As the risk of no-show decreases, risk and odds converge, becoming basically
  identical.
- We will expand on the utility of this next week when we introduce logistic
  regression.



Ratios
================================================================================

- ~Sometimes~ Often we want to compare the risk of two groups.
- For example, compare the risk of a no-show when the patient received a SMS
  reminder to the risk of a no-show when the patient did not receive a SMS
  reminder.


```{r}
# Saves my calculated results to risk_sms for the next step.

risk_sms <- 
  appointments |>
  group_by(received_sms) |>
  summarize(
    den = n(),
    num = sum(no_show),
    risk = 100*num/den
  )
risk_sms

```

- Appointments where the patient received a SMS message have a higher risk
  of no-show appointments than appointments where the patient did not receive a
  SMS reminder.
- Surely there's a way to express this as a single number!


RISK RATIO
--------------------------------------------------------------------------------

To calculate a risk ratio: 

- Divide the risk of the group of interest by the risk of the other group.
    - Yes. It is that simple.
- Risk Ratio:
    - A ratio greater than one indicates risk.
    - A ratio close to / equal to one indicates no risk.
    - A ratio smaller than one indicates protection.
- In public health, we tend to talk about risk factors a lot.
    - These are risk ratios GREATER than one.
- Social Workers like to also talk about protective features:
    - These are risk ratios LESS than one.
- Nobody is much interested in talking about ratios close to one, because they
  don't matter much.

```{r}
# Calculating Relative Risk
# Two paths to the same destination.

rr_1 <-
  risk_sms |>
  summarize(
    # Oh look, square brackets!!!
    risk_ratio = risk[received_sms == TRUE] / risk[received_sms == FALSE]
  )

rr_2 <-
  risk_sms |>
  summarize(
    risk_ratio = risk[received_sms] / risk[!received_sms]
  )

wrap_table(rr_1) / wrap_table(rr_2)
```


Odds Ratio
--------------------------------------------------------------------------------

We rarely need to use square bracket notation when using the tidyverse
functions, but I have never found another way to calculate risk and odds ratios.

Speaking of risk, let's calculate the Odds Ratio of no-show.

- Odds Ratio:
    - A ratio greater than one indicates risk.
    - A ratio close to / equal to one indicates no risk.
    - A ratio smaller than one indicates protection.


```{r}

# Getting fancy! All in one shot!
appointments |>
  group_by(received_sms) |>
  summarize(
    den = sum(!no_show),
    num = sum(no_show),
    odds = num/den
  ) |>
  summarize(
    odds_ratio = odds[received_sms]/odds[!received_sms]
  )

```

- Because no-show is not rare, the odds ratio is more dramatic than the risk
  ratio.
- It isn't wrong, we just don't easily think about things this way.
    - But it can be misleading.
- So why do we use odds at all?
    - Odds and risk converge for rare events.
    - Odds care conveniently connected to logistic regression.
    - But that's a different lecture.
